Hierarchical Principal Manifold Estimation

### 2023-10-02 paper/main.tex
Add changes made to oral qualifier proposal

\section*{Abstract}

@@

Neurodegenerative diseases such as Alzheimer's disease (AD) and Parkinson's disease are highly complex, chronic conditions that exhibit substantial heterogeneity between patients. To enable researchers to identify potential biomarkers of progression for these conditions, large multi-site longitudinal observational studies have been conducted that collect information from a wide range of participants over long durations across several modalities, including neuroimaging [Cite ADNI and PPMI studies]. Neuroimaging data in particular have been shown to be sensitive to developments in the early stages of disease progression, often before diagnosis occurs [Find citation from reading].

While observations from neuroimaging data are typically high-dimensional, it is commonly assumed that the data lie along a low-dimensional manifold [Relevant citations demonstrating this]. Under this assumption, manifold learning algorithms can be used to recover this underlying manifold and parameterize this high-dimensional input data in the learned low-dimensional space. Most commonly-used manifold learning algorithms, including Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), Locally Linear Embedding (\cite{roweisNonlinearDimensionalityReduction2000}), Laplacian Eigenmaps (\cite{belkinLaplacianEigenmapsDimensionality2003}), and Diffusion Maps (\cite{coifmanDiffusionMaps2006}) construct a graph of similarities between observations, which is then used to determine the structure of the manifold. \cite{mengPrincipalManifoldEstimation2021} present an alternative principal manifold framework for manifold estimation which extends to higher dimensions the principal curve framework introduced by \cite{hastiePrincipalCurves1989}, in which principal curves are defined as one-dimensional curves that pass through the middle of a high-dimensional dataset.

In neuroimaging settings, manifold learning algorithms are most often used to estimate the underlying manifold at the population level, with vectorized image intensities being used as the input data [Citations]. \cite{yueParameterizationWhiteMatter2016} used manifold learning to estimate a low-dimensional parameterization of the corpus callosum at the individual level. However, in many cases, the outcome of interest is framed as a comparison between different groups of individuals (e.g. treatment groups in a clinical trial). This goal is particularly relevant in the case of the complex neurodegenerative conditions mentioned above, where the high degree of heterogeneity in patient experiences has led to attempts to identify subgroups of patients based on a range of criteria such as symptom expression, onset, or rate of progression [Citations]. Fundamentally, if a low dimensional manifold is thought to effectively represent data related to the patient experience of a neurodegenerative condition, and the experience of that condition differs by group, then the low-dimensional representation of that experience should also differ by group.

@@

\section{Application to Simulated and MNIST Data}

\subsection{Application to Simulated Data}

\subsection{Application to MNIST Data}

\section{Application to ADNI Data}

\section{Discussion}

\section{Aim 2: Principal Manifold Estimation for Hierarchical Data} \label{sec:hpme}

\subsection{Background}

\subsection{Methods}\label{s:methods}

\subsubsection{Additive Smoothing Spline Model}

\subsubsection{Hierarchical Principal Manifold Estimation}

\subsection{Proposed Work}
Further work will be dedicated to testing the proposed approach in several settings. First, the algorithm will be tested with a simplified additive model on the MNIST dataset of handwritten images. This will provide an opportunity to test the method on a relatively small imaging dataset, serving as a proof of concept and indicating where problems may arise when the algorithm is used with more complex data structures. The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset. In both of these cases, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared. Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset. Specifically, the algorithm can be used to estimate the manifolds underlying the hippocampus of study participants. Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.

>>>>

\section{Abstract} \label{s:abstract}

@@

@@

\section{Proposed Work}
Further work will be dedicated to testing the proposed approach in several settings. First, the algorithm will be tested with a simplified additive model on the MNIST dataset of handwritten images. This will provide an opportunity to test the method on a relatively small imaging dataset, serving as a proof of concept and indicating where problems may arise when the algorithm is used with more complex data structures. The method will next be applied to a series of simulated datasets that were generated to reflect the hierarchical structure present in the ADNI dataset. In both of these cases, the estimates of the hierarchical approach will be compared to alternative methods, with the fit of each method to the data being compared. Finally, the hierarchical approach will be applied to structural MRI data from the ADNI dataset. Specifically, the algorithm can be used to estimate the manifolds underlying the hippocampus of study participants. Of interest in this case will be a comparison of the varying-coefficient models by group, which may provide a visual confirmation that the algorithm's estimates are detecting the most relevant information included in the data.
